Labeling in the Strucplots Framework

### 2005-09-18 trunk/vcd/inst/doc/labeling.Rnw
- vignettes revised - bugs in structable indexing fixed - mosaic, sieve, and assoc made structable aware - panel_generator class -> generating_function - offset_labels and offset_varnames added to labeling_border - spacing_highlighting replaces spacing_doubledecker
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@457 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\var}[1]{\textit{\texttt{#1}}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}

@@

\title{Labeling in Strucplots}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\email{David.Meyer@R-project.org}
}
\maketitle
\sloppy

One of the major enhancements in \pkg{vcd} compared to
\codefun{mosaicplot} and \codefun{assocplot} in
base {\sf R} is the labeling in the strucplot framework which now
has become much more powerful and flexible. Like the panel, the
shading, the spacing, and the legend, labeling is now completely modular.
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \codefun{strucplot} which then draws the labels (generating
functions are discussed in a separate vignette). The labeling is 
well-separated from the actual plotting that occurs in the panel functions;
it actually only relies on the viewport tree produced by them, and the
`dimnames' attribute of the visualized table. The drawing of the
labels happens after the actual plot has been drawn by the panel function.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
% As an example, consider \codefun{labeling\_list}:

% \begin{Sinput}
% labeling_list <- function(gp = gpar(), just = "left", pos = "left", lsep = ": ", 
%                           sep = " ", offset = unit(c(2, 2), "lines"),
%                           varnames = TRUE, cols = 2, ...) 
%   function(d, split_vertical, condvars) {
%     \dots
%   }
% \end{Sinput}

% \codefun{labeling\_list} is a generating function that takes all parameters
% as arguments, and returns the actual labeling function which takes
% only three arguments: the dimnames structure of the table, the logical
% vector with the splitting information, and the number of conditioning
% variables (if any). The labeling function has access to the arguments
% of the generating function because in {\sf R}, nested functions are
% lexically scoped. 
In the following, we describe the three basic modules
(\codefun{labeling\_text}, \codefun{labeling\_list}, and \codefun{labeling\_cells}) and
derived functions that build upon them.

@@

As an example, consider the `Titanic' data, consisting
of 4 categorical variables: Survived, Sex (i.e., `Gender'), Age, and
Crew. By default, the variable names and levels are plotted `around' the
plot in a counter-clockwise way (see Figure \ref{fig:defaults}):

@@

\begin{figure}[p]
\begin{center}
<<defaultfig,fig=TRUE,echo=FALSE>>=
<<default>>
@
\caption{Mosaic plot for the Titanic data.}
\label{fig:defaults}
\end{center}
\end{figure}

@@

\noindent The \code{clip} parameter is passed to the labeling function
via the \code{labeling\_args} argument which takes a list of
parameters. \code{clip} itself takes a vector of logicals (one for
each dimension). Almost all vectorized arguments in the strucplot
framework can be abbreviated in the following way: unnamed components
(or the defaults, if there are none) are recycled as needed, but
overridden by the named components. Here, the default is \code{FALSE},
and therefore clipping is enabled only for the `Survived' and `Age' dimensions.
A more sensible solution to the overlap problem is to abbreviate the 
levels (see Figure \ref{fig:abbreviating}):

@@

<<rotate>>=
mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0)),
                margins = c(right = 6, bottom = 4))
@

@@

\noindent \code{labbl\_varnames} (``variable names to the bottom/left of the
labels'') is a vector of logicals indicating the side for the
variable names. The resulting layout is basically what \codefun{labeling\_lboxed} produces,
and a similar design is used by the \codefun{doubledecker}.

@@

This labeling draws both variable names and levels in the
cells. As an example, we use the `PreSex' data on pre- and
extramarital sex and divorce (see Figure \ref{fig:cell}):

@@

\begin{figure}[h]
\begin{center}
<<cellfig,fig=TRUE,echo=FALSE>>=
<<cell>>
@
\caption{Cell labeling for the `PreSex' data.}
\label{fig:cell}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
<<cell2fig,fig=TRUE,echo=FALSE>>=
<<cell2>>
@
\caption{Cell labeling for the `PreSex' data, labels abbreviated.}
\label{fig:cell2}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
<<conditionalfig,fig=TRUE,echo=FALSE>>=
<<conditional>>
@
\caption{Conditional labeling for the `PreSex' data, labels abbreviated.}
\label{fig:conditional}
\end{center}
\end{figure}

\noindent Additionally, the cell labeling allows the user to add
arbitrary text to the cells by supplying a character array in the same
shape than the data array to the \code{text} argument (cells with missing values
are ignored). In the following example, this is used to add all observed values greater
than 5 to the cells after the mosaic for the `Titanic' data has been plotted
(see Figure \ref{fig:text}):

<<text>>=
mosaic(Titanic, pop = FALSE, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)))

@@

\begin{figure}[h]
\begin{center}
<<textfig,fig=TRUE,echo=FALSE>>=
<<text>>
@
\caption{User-supplied Text added to a mosaic display of the `Titanic' data.}
\label{fig:text}
\end{center}
\end{figure}

\section{A simple list of labels: \texttt{labels\_list()}}

>>>>

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\var}[1]{\textit{\texttt{#1}}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
\newcommand{\data}[1]{`\texttt{#1}'}

@@

\title{Labeling in the Strucplots Framework}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\email{David.Meyer@R-project.org}
}
\maketitle
\sloppy

One of the major enhancements in package \pkg{vcd} compared to
\codefun{mosaicplot} and \codefun{assocplot} in
base {\sf R} is the labeling in the strucplot framework which 
offers much more features and flexibility. Like the core, 
shading, spacing, and legend modules, labeling is now completely
modular (see the introducing vignette for an overview on
the strucplot framework).
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \codefun{strucplot} which then draws the labels (the idea behind generating
functions is discussed in detail in the vignette on colors and shadings). 
Labeling is well-separated from the actual plotting that occurs in the
low-level core functions. It only relies on the viewport tree produced by them, and the
`\texttt{dimnames}' attribute of the visualized table. The drawing of the
labels happens after the actual plot has been drawn by the core function.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
In the following, we describe the three basic modules
(\codefun{labeling\_text}, \codefun{labeling\_list}, and \codefun{labeling\_cells}) and
derived functions that build upon them.

@@

As an example, consider the \data{Titanic} data, consisting
of 4 categorical variables: Survived, Sex (i.e., `Gender'), Age, and
Crew. By default, the variable names and levels are plotted `around' the
plot in a counter-clockwise way (see Figure \ref{fig:defaults}):

@@

\begin{figure}[p]
\begin{center}
<<defaultfig,fig=TRUE,echo=FALSE>>=
<<default>>
@
\caption{Mosaic plot for the \data{Titanic} data.}
\label{fig:defaults}
\end{center}
\end{figure}

@@

\noindent The \code{clip} parameter is passed to the labeling function
via the \code{labeling\_args} argument which takes a list of
parameters. \code{clip} itself takes a vector of logicals (one for
each dimension). 
% as mentioned before
Almost all vectorized arguments in the strucplot
framework can be abbreviated in the following way: unnamed components
(or the defaults, if there are none) are recycled as needed, but
overridden by the named components. Here, the default is \code{FALSE},
and therefore clipping is enabled only for the `Survived' and `Age' dimensions.
A more sensible solution to the overlap problem is to abbreviate the 
levels (see Figure \ref{fig:abbreviating}):

@@

<<rotate>>=
mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), margins = c(right = 4, bottom = 3))
@

@@

\noindent \code{labbl\_varnames} (``variable names to the bottom/left of the
labels'') is a vector of logicals indicating the side for the
variable names. The resulting layout is basically what \codefun{labeling\_lboxed} produces,
and a similar design is used by the \codefun{doubledecker} function.

@@

This labeling draws both variable names and levels in the
cells. As an example, we use the \data{PreSex} data on pre- and
extramarital sex and divorce (see Figure \ref{fig:cell}):

@@

\begin{figure}[h]
\begin{center}
<<cellfig,fig=TRUE,echo=FALSE>>=
<<cell>>
@
\caption{Cell labeling for the \data{PreSex} data.}
\label{fig:cell}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
<<cell2fig,fig=TRUE,echo=FALSE>>=
<<cell2>>
@
\caption{Cell labeling for the \data{PreSex} data, labels abbreviated.}
\label{fig:cell2}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
<<conditionalfig,fig=TRUE,echo=FALSE>>=
<<conditional>>
@
\caption{Conditional labeling for the \data{PreSex}, labels abbreviated.}
\label{fig:conditional}
\end{center}
\end{figure}

\noindent Additionally, the cell labeling allows the user to add
arbitrary text to the cells by supplying a character array in the same
shape than the data array to the \code{text} argument (cells with missing values
are ignored). In the following example using the `Titanic' data, 
this is used to add all observed values greater
than 5 to the cells after the mosaic has been plotted (see Figure \ref{fig:text}):

<<text>>=
mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)), pop = FALSE)

@@

\begin{figure}[h]
\begin{center}
<<textfig,fig=TRUE,echo=FALSE>>=
<<text>>
@
\caption{User-supplied Text added to a mosaic display of the \data{Titanic} data.}
\label{fig:text}
\end{center}
\end{figure}

\section{A simple list of labels: \texttt{labeling\_list()}}
